---
title: "Predicting Future Scooter Movement - Poisson Regression"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introdcution 
We want to predict and model the position of the scooters position in the future. 

We estimate that the scooter rides starting in a given cell of the grid follows a poission distribution. That is linearly dependent on the cells poistion, the hour of the day and the day of the week. (Later one can add other covariates such as weather, and month)

We start by fitting a generelized linear model based on the possion family.

```{r}
library(dplyr)
library("rgdal")
library("sf")
library(ggmap)
library(lubridate)

# Reading data
df <- read.csv("processed-data/output10min.csv")

# Removing index columns
df <- df[,-c(1,2)]
# Add columns about time stamps
df$time <- as.POSIXct(df$time, "%Y-%m-%d %H:%M:%S", tz="Europe/Paris")

# Finding start and end of trips
df.trip <- df %>% group_by(operator, id) %>% mutate(isEnd = ifelse((lead(distanceTravelled)<50 | lead(distanceTravelled)>3000) & (distanceTravelled > 50 | distanceTravelled > 3000),1,0),
                                               isStart = ifelse((distanceTravelled < 50 | distanceTravelled > 3000) & (lead(distanceTravelled) > 50 & lead(distanceTravelled) < 3000),1,0)) %>% ungroup()

# Finding intervals part of a trip inbetween start and end. 
df.trip <- df.trip %>% group_by(operator, id) %>% mutate(isMid = ifelse((lead(isStart) == 1 | (lead(distanceTravelled)> 50 & lead(distanceTravelled) < 3000)) & (lag(distanceTravelled)> 50 & lag(distanceTravelled) < 3000),1,0)) %>% ungroup()

df.trip$monthDay <- strftime(df.trip$time, format = "%m/%d")
df.trip$weekday <- strftime(df.trip$time, format = "%a")
df.trip$hour <- strftime(df.trip$time, format = "%H")
df.trip$month <- strftime(df.trip$time, format = "%m")
df.trip$year <- strftime(df.trip$time, format = "%Y")
df.trip$day <- strftime(df.trip$time, format = "%d")
df.trip$min <- strftime(df.trip$time, format = "%M")


df.trip <- df.trip %>%
  st_as_sf(coords = c("lon", "lat"), crs = "+proj=longlat +zone=33") %>%
  st_transform(crs="+proj=utm +zone=33") %>%st_crop(xmin = 254328.471, xmax = 268685.126, ymin = 6646893.804, ymax = 6655335.032)

df.start <- df.trip[df.trip$isStart == 1,]
df.start <- df.start[!is.na(df.start$time),]
```


Make grid:
```{r}
coast <- st_read('gis/shapeFiles/land_polygons.shp')
coast.c <- st_crop(coast, xmin = 10.54433, xmax = 10.92609, ymin = 59.88134, ymax = 59.98148)
coast.c <- st_transform(coast.c, CRS("+proj=utm +zone=33 ellps=WGS84"))

grid_cell_size = c(300, 300) # In meters
grid <- st_make_grid(coast.c, cellsize = grid_cell_size)
grid <- st_intersection(coast.c, grid)
grid$grid_id <- 1:nrow(grid)
grid <- grid[,c("grid_id","geometry")]
grid %>%
  write_sf("predict_grid_300.shp")

```

Start rate parameter:
```{r}
df.start <- df.start  %>% st_join(grid, join = st_intersects) %>% as.data.frame()
df.start$weekday <- strftime(df.start$time, format = "%a")

df.start_count <- df.start %>% group_by(grid_id, hour, weekday, year, month, day) %>% summarise(started_trips = n())
df.start_count <- na.omit(df.start_count)

```

Where does trips go? 
```{r}
df.trip <- df.trip %>% st_join(grid, join = st_intersects) %>% as.data.frame()

## Where does the trips go?
df.start_end <- df.trip[df.trip$isStart == 1 | df.trip$isEnd == 1,]
df.start_end <- df.start_end[!is.na(df.start_end$time),]

df.start_end <- df.start_end %>% group_by(operator, id) %>% mutate(start_grid = grid_id, end_grid = lead(grid_id))
df.start_end <- df.start_end[df.start_end$isStart == 1, ]
df.start_end <- df.start_end %>% group_by(start_grid, end_grid) %>% summarise(n=n()) # if more data can use hour and weekday here
df.start_end <- na.omit(df.start_end)
df.start_end <- df.start_end %>% group_by(start_grid) %>% mutate(p = n/n())
df.start_end<- df.start_end[seq(dim(df.start_end)[1],1),] # Reverse order


## How many scooter in area at the start of an hour
df.grid_count <- df.trip[df.trip$min < 10,]
df.grid_count <- df.trip %>% group_by(grid_id, hour, weekday, day, month, year) %>% summarise(number_scooters_at_hr_start = n_distinct(id))
df.grid_count <- na.omit(df.grid_count)

###
df.grid_movment <- left_join(df.start_count, df.grid_count, by=c("grid_id", "hour", "weekday", "day", "month", "year"))

df.grid_movment$time <- as.POSIXct(paste(df.grid_movment$year, df.grid_movment$month, df.grid_movment$day, df.grid_movment$hour), "%Y %m %d %H", tz="Europe/Paris")
df.grid_movment$predicted <- 0

max_time <- max(df.grid_movment$time)
start_year = strftime(max_time, format = "%Y")
start_hr = strftime(max_time, format = "%H")
start_day = strftime(max_time, format = "%d")
start_month = strftime(max_time, format = "%m")

df.predict_from = df.grid_movment[df.grid_movment$hour == start_hr &
                                    df.grid_movment$year == start_year &
                                    df.grid_movment$day == start_day &
                                    df.grid_movment$month == start_month,]
#grid_ids <- 1:nrow(grid)
#grid_ids_to_add <- as.data.frame(grid_ids[!grid_ids %in% df.predict_from$grid_id])
#names(grid_ids_to_add) <- c("grid_id")
#grid_ids_to_add$time <- max_time
#grid_ids_to_add$weekday <- strftime(grid_ids_to_add$time, format = "%a")
#grid_ids_to_add$hour <- strftime(grid_ids_to_add$time, format = "%H")
#grid_ids_to_add$month <- strftime(grid_ids_to_add$time, format = "%m")
#grid_ids_to_add$year <- strftime(grid_ids_to_add$time, format = "%Y")
#grid_ids_to_add$day <- strftime(grid_ids_to_add$time, format = "%d")
#grid_ids_to_add$number_scooters_at_hr_start <- 0 
#grid_ids_to_add$predicted <- 0
#grid_ids_to_add <- grid_ids_to_add[, c("grid_id", "weekday", "hour", "year", "month", "day", "time", #"number_scooters_at_hr_start", "predicted")] 
#grid_ids_to_add$started_trips <- 0
#df.grid_movment <- bind_rows(df.grid_movment, grid_ids_to_add)

df.grid_movment$grid_id <- as.factor(df.grid_movment$grid_id)

```
Test and train data:
```{r}
max_time <- max(df.grid_movment$time)
train.max_time <- max_time - 60*60*20 # 12 hr
df.train <- df.grid_movment[df.grid_movment$time < train.max_time, ]
df.test <- df.grid_movment[df.grid_movment$time >= train.max_time, ]
```

Fit glm to estimate 
```{r}
model <- glm(started_trips ~ grid_id + hour + weekday + number_scooters_at_hr_start, poisson, df.train)
summary(model)
```

Overdispersion:
```{r}
overdispersionparameter = model$deviance / model$df.residual
overdispersionparameter

p.overdisp <- pchisq(model$deviance, df=model$df.residual, lower.tail = F)
p.overdisp
```

```{r}
model <- glm(started_trips ~ grid_id + hour + weekday + number_scooters_at_hr_start, quasipoisson(), df.train)
summary(model)

overdispersionparameter = model$deviance / model$df.residual
```
Simulating:
```{r}
predict_movement <- function(df, hours_into_future){
  df$predicted_rate <- predict.glm(model, df, type="response")*hours_into_future 
  df$started_rides <- rpois(nrow(df), df$predicted_rate) 
  df <- na.omit(df)
  
  movement_probs <- runif(sum(df$started_rides), max = 1.2)
  grids <- unique(df.grid_movment$grid_id)
  count <- 0
  
  df_new <- as.data.frame(unique(df.grid_movment$grid_id), colnames = c("grid_id"))
  names(df_new) <- c("grid_id")
  df_new$time <- df$time[1]+ hours_into_future*60*60
  df_new$weekday <- strftime(df_new$time, format = "%a")
  df_new$hour <- strftime(df_new$time, format = "%H")
  df_new$month <- strftime(df_new$time, format = "%m")
  df_new$year <- strftime(df_new$time, format = "%Y")
  df_new$day <- strftime(df_new$time, format = "%d")
  df_new <- left_join(df_new, df[,c("grid_id", "number_scooters_at_hr_start","started_rides")], by=c("grid_id"))
  df_new[is.na(df_new)] <- 0
  df_new$scooters_left <- df_new$number_scooters_at_hr_start - df_new$started_rides
  df_new[df_new$scooters_left < 0, ]$scooters_left <- 0 
  df_new$started_rides <- df_new$number_scooters_at_hr_start - df_new$scooters_left
  df_new$scooters_arrived <- 0 

  for(i in 1:nrow(df_new)){
    print(i)
    if(df_new$started_rides[i] > 0){
      for(j in 1:df_new[i,]$started_rides){
        count <- count + 1
        from_grid = df[i,]$grid_id
        move_prob <- movement_probs[j]
        
        end_grid <- NA
        if(move_prob < 1){
          end_grid = df.start_end[df.start_end$start_grid == from_grid & df.start_end$p >= move_prob, ]$end_grid[1]
        }
        
        if(is.na(end_grid)){
          end_grid = sample(grids, 1)
        }
        
        df_new[df_new$grid_id == end_grid,]$scooters_arrived = df_new[df_new$grid_id == end_grid,]$scooters_arrived + 1
           
      }
    }
    
  }
  
  df_new$number_scooters_at_hr_start <- df_new$scooters_left + df_new$scooters_arrived

  df_new$predicted <- 1
  df_new <- df_new[, c("grid_id", "hour", "weekday", "year", "month", "day", "time", "number_scooters_at_hr_start", "predicted")]
  
  return(df_new)
}


add_predicted_data <- function(df, n, time_step_hr){
  for(i in 1:n){
    print(i)
    max_time <- max(df$time)
    start_year = strftime(max_time, format = "%Y")
    start_hr = strftime(max_time, format = "%H")
    start_day = strftime(max_time, format = "%d")
    start_month = strftime(max_time, format = "%m")
    
    df.predict_from = df[df$hour == start_hr &
                                        df$year == start_year &
                                        df$day == start_day &
                                        df$month == start_month,]
    

    new_data = predict_movement(df.predict_from, time_step_hr)
    new_data <- na.omit(new_data)
    
    df = bind_rows(df, new_data)
    print(nrow(new_data))
    df <- na.omit(df)
   
  }
  return(df)

}

df.train <- df.train[,-c(7)]
df.train$predicted <- 0
df.train <- df.train[, c("grid_id", "hour", "weekday", "year", "month", "day", "time", "number_scooters_at_hr_start", "predicted")]
res <- c()
for(k in 1:100){
  print(k)
  df.result = add_predicted_data(df.train, 1, 1)
  df.result <-df.result[df.result$predicted == 1, ] 
  df.result <- na.omit(df.result)
  df.result <- full_join(df.test, df.result, by=c("grid_id","hour", "year", "month", "day"))
  df.result <- na.omit(df.result)
  df.result[is.na(df.result)] <- 0
  df.result$err <- df.result$number_scooters_at_hr_start.x - df.result$number_scooters_at_hr_start.y
  mse_1hr = mean((df.result$err)^2)**0.5
  res <- c(res, mse_1hr)
}

#Write results
#df.result$grid_id <- as.integer(df.result$grid_id)
#df.predict_output <- df.result %>% left_join(grid, by=c("grid_id"))
#write.csv(df.predict_output, "predicted.csv")

## Baseline same as day before
df.baseline <- df.train 
df.baseline$time <- df.baseline$time + 24*60*60
df.baseline$monthDay <- strftime(df.baseline$time, format = "%m/%d")
df.baseline$weekday <- strftime(df.baseline$time, format = "%a")
df.baseline$hour <- strftime(df.baseline$time, format = "%H")
df.baseline$month <- strftime(df.baseline$time, format = "%m")
df.baseline$year <- strftime(df.baseline$time, format = "%Y")
df.baseline$day <- strftime(df.baseline$time, format = "%d")
df.baseline$min <- strftime(df.baseline$time, format = "%M")
df.baseline <- left_join(df.test, df.baseline, by=c("grid_id","hour", "year", "month", "day"))
df.baseline <- na.omit(df.baseline)
df.baseline$err <- df.baseline$number_scooters_at_hr_start.x - df.baseline$number_scooters_at_hr_start.y
mse = mean((df.baseline$err)^2)**0.5



```



